---
title: "Group 2 — Risk factors"
subtitle: "Studio wall"
author: "Siva, Shaariq and Taniska"
format:
  html:
    toc: true
    toc-depth: 2
code-tools: true
---

# Spine

> How do bodies carry risk, how do class and smoke talk, and where do children still die young?

# Introduction — the tables we own

This folder is ours for the ten days. Three tables, three shapes.

- **Panel A** (`panel-A.csv`) — `cardioRiskFactors`: BMI, lipids, blood pressure, uric acid. `sex` and `smok` have already been recoded from integers to factors.
- **Panel B** (`panel-B.csv`) — Scottish Health Survey 1998: sex, social class, smoking, diagnosed CVD. `area` is a code, not a map.
- **Panel C** (`panel-C.csv`) — SOWC child / infant / neonatal mortality by country. Join by country name.

If a brick needs a histogram, it lives in A. If it needs a mosaic, it lives in B.

```{r}
#| label: setup
#| message: false # keeps rendered Quarto less noisy

library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)

A <- read_csv("data/panel-A.csv")
B <- read_csv("data/panel-B.csv")
```

# Panel A — body / measure

> *bmi, choles, sys, dia, Uric are Quant. Do not average `smok`.*

### 1. See the table

```{r}
#| label: inspect-panel-A-data-1

# glimpse() / inspect() / skim()
dplyr::glimpse(A)
```

Our Panel-A data has:

- 998 observations and 14 variables. One case = one person, measured once.
- `age`: Quantitative, numeric
- `bmi`: Quantitative, numeric
- `waisthip`: Quantitative, numeric
- `smok`: Qualitative, categorical. Encoded `<chr>`, to be mutated
- `choles`: Quantitative, numeric
- `trig`: Quantitative, numeric
- `hdl`: Quantitative, numeric
- `ldl`: Quantitative, numeric
- `sys`: Quantitative, numeric
- `dia`: Quantitative, numeric
- `Uric`: Quantitative, numeric
- `sex`: Qualitative, categorical. Encoded `<chr>`, to be mutated
- `alco`: Quantitative, numeric
- `apoa`: Quantitative, numeric

```{r}
#| label: point-at-a-row

A |> slice(1)
```

Row 1 is one person: a 52-year-old woman who doesn't smoke or drink, with a BMI of 36.9 and blood pressure of 154/89.

```{r}
#| label: inspect-panel-A-data-2

# glimpse() / inspect() / skim()
skimr::skim(A)
```

On the Qual variables, panel-A data has:

- `sex` has 2 levels: "female" (524) and "male" (474)
- `smok` has 2 levels: "non_smoker" (775) and "smoker" (223)
- The split inside `sex` is very uneven: only **36 women smoke** (vs 488 who don't), while **187 men smoke** (vs 287 who don't)

```{r}
#| label: qual-counts

mosaic::tally(~ sex, data = A)
mosaic::tally(~ smok, data = A)
mosaic::tally(smok ~ sex, data = A) # counts of smok inside each sex
```

On the Quant variables, panel-A data has:

- `age`: mean 44.6, sd 11.5, median 45, min 25, max 64. Pretty flat, no skew.
- `bmi`: mean 25.8, sd 5.3, median 25.1, min 14.8, max 48.6. Skewed right.
- `hdl`: mean 1.54, sd 0.45, median 1.48. Slightly skewed right.
- `ldl`: mean 3.45, sd 1.22, median 3.42. Close to symmetric.
- `sys`: mean 135.9, sd 23.9, median 132, max 245. Skewed right.
- `Uric`, `trig`, `alco`: very skewed right with a long tail (`Uric` goes up to 1885).
- `alco`: **721 of 998 people have 0**, so most people don't drink at all.

There is no missing data.

```{r}
#| label: mutate-A-factors

A_mod <- A |> mutate(sex = as.factor(sex), smok = as.factor(smok))

glimpse(A_mod)
```

A few helper columns for later bricks. These are **cut-offs we chose**, not columns in the data: "high bp" is `sys >= 140` or `dia >= 90`, "obese" is `bmi >= 30`, "drinker" is `alco > 0`.

```{r}
#| label: helper-columns

A_flag <- A_mod |>
  mutate(
    high_bp = (sys >= 140 | dia >= 90),
    obese   = (bmi >= 30),
    drinker = (alco > 0)
  )

A_flag |> summarise(
  pct_high_bp = mean(high_bp),
  pct_obese   = mean(obese),
  n_nondrinkers = sum(!drinker)
)
```

Findings from this look:

1.  524 women and 474 men; 775 non-smokers and 223 smokers (only 36 women smoke, compared to 187 men).
2.  Most people don't drink: 721 of 998.
3.  48% of people have high blood pressure.
4.  20% of people are obese (`bmi >= 30`).

### 2. Ask

Based on the spine, these are our questions:

Possibly can answer:

1.  Are women and men affected differently? Do they carry risk in different ways?
2.  Is `waisthip` a better match for risk markers than `bmi`? (Where you carry weight vs how much.)
3.  Do smokers and non-smokers have different `hdl` and `ldl` levels, for men and for women? Is the pattern still there for non-drinkers? (Main Character Question)

Possibly cannot answer:

1.  How much smoking matters. (No dose or duration of smoking, just smoker / non-smoker.)
2.  Did any of these people get heart disease? (No outcome column, so we can't tell whether a "high" value actually leads to heart disease.)

### 3. Choose a mark

1.  Bar chart for the Qual counts (`sex`, `smok`), no histograms for these.
2.  Histograms for single Quant variables (`age`, `bmi`, `sys`, `alco`, ...)
3.  Scatter + straight-line smoother for Quant + Quant (`sys ~ age`, `sys ~ waisthip`, `sys ~ bmi`, `hdl ~ trig`, `hdl ~ alco`), with a correlation number to back it up
4.  Violins for Quant + Qual (2 levels): body measures by `sex`, and `hdl` / `ldl` by `smok`, split by `sex`
5.  Density plot to see `hdl` and `ldl` side by side in one chart
6.  Two-sample test for means beside the matching violin

### 4. Predict

1.  **Question 1:** Men will carry more weight around the middle (`waisthip`), even if women have the higher `bmi`.
2.  **Question 2:** `waisthip` will line up with blood pressure and uric acid more closely than `bmi` does.
3.  **Question 3 (Main Character):** Smokers will have worse cholesterol: lower `hdl` and higher `ldl` than non-smokers.

### 5. Make

These are the charts we used to explore the data. They don't show up in the render, but they helped us decide what to look at next.

```{r}
#| label: panelA-charts-I-want-to-see-but-not-show-1
#| eval: false
#| echo: true # Include the code, but don't compute in render

## Qual counts
A_mod |> gf_bar(~ sex)
A_mod |> gf_bar(~ smok)
A_mod |> gf_bar(~ smok, fill = ~ sex)

## Quant shapes
A_mod |> gf_histogram(~ age)
A_mod |> gf_histogram(~ bmi)
A_mod |> gf_histogram(~ sys)
A_mod |> gf_histogram(~ Uric)
A_mod |> gf_histogram(~ alco)
```

**Question 1: do women and men carry risk differently?**

```{r}
#| label: women-vs-men

A_mod |> gf_violin(bmi ~ sex, fill = ~ sex) |>
  gf_labs(title = "Women have the higher BMI...", x = NULL)
A_mod |> gf_violin(waisthip ~ sex, fill = ~ sex) |>
  gf_labs(title = "...but men carry more weight around the middle", x = NULL)
A_mod |> gf_violin(sys ~ sex, fill = ~ sex) |>
  gf_labs(title = "Men have higher blood pressure", x = NULL, y = "Systolic BP (mmHg)")
A_mod |> gf_violin(Uric ~ sex, fill = ~ sex) |>
  gf_labs(title = "Men have much higher uric acid", x = NULL, y = "Uric acid")
```

Prediction was right: women have higher `bmi`, men have higher `waisthip`, blood pressure and `Uric`.

Blood pressure rises with age:

```{r}
#| label: bp-age

A_mod |> gf_point(sys ~ age) |> gf_smooth(method = "lm") |>
  gf_labs(title = "Blood pressure goes up with age", y = "Systolic BP (mmHg)")
```

**Question 2: where you carry weight (`waisthip`) vs how much (`bmi`)**

```{r}
#| label: waisthip-vs-bmi

A_mod |> gf_point(sys ~ waisthip) |> gf_smooth(method = "lm")
A_mod |> gf_point(sys ~ bmi) |> gf_smooth(method = "lm")

A_mod |> gf_point(Uric ~ waisthip) |> gf_smooth(method = "lm")
A_mod |> gf_point(Uric ~ bmi) |> gf_smooth(method = "lm")

A_mod |> gf_point(trig ~ waisthip) |> gf_smooth(method = "lm")
A_mod |> gf_point(trig ~ bmi) |> gf_smooth(method = "lm")

# put numbers on it
cor(sys ~ waisthip, data = A_mod)
cor(sys ~ bmi, data = A_mod)
cor(Uric ~ waisthip, data = A_mod)
cor(Uric ~ bmi, data = A_mod)
cor(trig ~ waisthip, data = A_mod)
cor(trig ~ bmi, data = A_mod)
```

`waisthip` wins every time: `sys` 0.33 vs 0.21, `Uric` 0.49 vs 0.11, `trig` 0.39 vs 0.22.

More triglycerides, less HDL:

```{r}
#| label: trig-hdl

A_mod |> gf_point(hdl ~ trig) |> gf_smooth(method = "lm") |>
  gf_labs(title = "More triglycerides, less good cholesterol")
```

Drinkers have higher uric acid:

```{r}
#| label: drinkers-uric

A_flag |> gf_violin(Uric ~ drinker, fill = ~ drinker) |>
  gf_labs(title = "Drinkers have higher uric acid", x = "Drinks alcohol?")
A_flag |> group_by(drinker) |> summarise(mean_Uric = mean(Uric), n = n())
```

**Question 3 (Main Character): smokers and cholesterol, men and women separately**

Human brick first: our own starting chart, before any AI.

```{r}
#| label: panel-A-human-brick
#| eval: false
#| echo: true

# Our first attempt: HDL and LDL as two separate graphs
A_mod |> gf_histogram(~ hdl, fill = ~ smok, position = "dodge", bins = 25) |> gf_facet_wrap(~ factor(sex))
A_mod |> gf_histogram(~ ldl, fill = ~ smok, position = "dodge", bins = 25) |> gf_facet_wrap(~ factor(sex))
```

Problem: two separate graphs, so it's hard to compare HDL and LDL side by side.

```{r}
#| label: panel-A-ai-prompt
#| eval: false
#| echo: true

# ## Prompt we actually used
# "how can I plot these 2 graphs together so I can analyse both hdl and ldl side by side"

# ai_draft
A_mod |>
  pivot_longer(c(hdl, ldl), names_to = "cholesterol", values_to = "level") |>
  gf_density(~ level, fill = ~ smok, alpha = 0.5) |>
  gf_facet_grid(cholesterol ~ factor(sex))

# ours
# kept pivot_longer() + gf_facet_grid(), tried it as a histogram first,
# switched to density because the groups are really uneven (36 vs 488 women),
# added a title + caption
```

```{r}
#| label: panel-A-hdl-ldl-together

A_mod |>
  pivot_longer(c(hdl, ldl), names_to = "cholesterol", values_to = "level") |>
  gf_density(~ level, fill = ~ smok, alpha = 0.5) |>
  gf_facet_grid(cholesterol ~ sex) |>
  gf_labs(
    title = "Male smokers have more good cholesterol and less bad; women barely differ",
    caption = "Panel A, cardioRiskFactors (n = 998). Only 36 women smoke.",
    x = "Cholesterol (mmol/L)"
  )
```

The same comparison as violins, with the group means:

```{r}
#| label: panel-A-Main-Character-question

A_mod |> gf_violin(hdl ~ smok, fill = ~ smok) |> gf_facet_grid(~ sex) |>
  gf_labs(title = "Male smokers have higher good cholesterol (HDL)", x = NULL,
          caption = "Only 36 women smoke, so the female smoker violin is thin on data.")
A_mod |> gf_violin(ldl ~ smok, fill = ~ smok) |> gf_facet_grid(~ sex) |>
  gf_labs(title = "Male smokers have lower bad cholesterol (LDL)", x = NULL)

A_mod |> group_by(sex, smok) |>
  summarise(n = n(), mean_hdl = mean(hdl), mean_ldl = mean(ldl), .groups = "drop")
```

The male smokers stand out: higher `hdl`, lower `ldl`. That's the opposite of what we predicted. What else is different about them? Check drinking:

```{r}
#| label: male-smokers-drink

A_flag |> filter(sex == "male", smok == "smoker") |>
  summarise(pct_drinkers = mean(drinker))   # about 68%

A_mod |> gf_point(hdl ~ alco) |> gf_smooth(method = "lm") |>
  gf_labs(title = "Good cholesterol rises a little with alcohol")
```

`hdl` rises with `alco`. So is it the smoking, or the drinking riding along with it? Repeat the comparison for **non-drinkers only**:

```{r}
#| label: non-drinkers-only

A_nd <- A_flag |> filter(!drinker)

A_nd |> gf_violin(hdl ~ smok, fill = ~ smok) |> gf_facet_grid(~ sex) |>
  gf_labs(title = "Non-drinkers only: male smokers still have higher HDL", x = NULL)
A_nd |> gf_violin(ldl ~ smok, fill = ~ smok) |> gf_facet_grid(~ sex) |>
  gf_labs(title = "Non-drinkers only: male smokers still have lower LDL", x = NULL)
```

Stat test for means, beside the matching violin (men only, since the women smoker group is just 36 people):

```{r}
#| label: panel-A-stat-test

A_men <- A_mod |> filter(sex == "male")

mosaic::t.test(hdl ~ smok, data = A_men)
mosaic::t.test(ldl ~ smok, data = A_men)

# and again with non-drinking men only
A_men_nd <- A_men |> filter(alco == 0)

mosaic::t.test(hdl ~ smok, data = A_men_nd)
mosaic::t.test(ldl ~ smok, data = A_men_nd)
```

### 6. Say

1.  Prediction missed. Male smokers have **higher** `hdl` (about 0.24 higher on average) and **lower** `ldl` (about 0.5 lower) than male non-smokers, both p-values far below 0.05.
2.  The pattern is still there among men who don't drink (`hdl` about 0.23 higher, `ldl` about 0.5 lower), so drinking alone doesn't explain it. Among women the gap is small, and only 36 women smoke.
3.  Women carry risk differently from men: higher `bmi` in women; higher `waisthip`, blood pressure and `Uric` in men.
4.  Blood pressure rises with age, and `waisthip` lines up with the risk markers (`Uric`, `trig`, `sys`) more closely than `bmi` does.
5.  48% of this group has high blood pressure, 20% is obese, and 721 of 998 don't drink.

### 7. Limit

- We can't say how much smoking matters: we only have smoker / non-smoker, no amount or duration.
- We can't say whether anyone got heart disease. There's no outcome column, so a "high" value can't be turned into "more likely to get heart disease".
- "High blood pressure", "obese" and "drinker" are cut-offs we chose, not columns in the data.
- This is one sample of 998 people. Associations here are not causes: a smoker's higher `hdl` is not a reason to smoke.
- Only 36 women smoke, too few to be sure about women.
- We don't know who these people are, where they're from, or how they were chosen.
- `ldl` is calculated from `choles`, `hdl` and `trig`, not measured on its own.
- Some values are extreme (`dia` up to 172, `Uric` up to 1885).
- Kit table we refused to abuse: `smok` and `sex` were never averaged or put in a histogram.

```{r}
A_mod |> gf_histogram(~ choles, fill = ~ sex)

A_mod |> group_by(sex) |> summarize ( mean_choles = mean(choles, na.rm = T))
A_mod |> group_by(smok) |> summarize ( mean_choles = mean(choles, na.rm = T))

mosaic::t_test(choles ~ sex, data = A_mod)
```

# Panel B — category / talk

> *smoke × sex, cvd × sc (social class). `area` is forbidden as geography. Do not average `sc`.*

### 1. See the table

```{r}
#| label: inspect-panel-B-data-1

# glimpse() / inspect() / skim()
dplyr::glimpse(B)
```

Our Panel-B data has:

- 8804 observations and 7 variables. One case = one person in the Scottish Health Survey, 1998.
- `age`: Quantitative, numeric
- `sex`: Qualitative, categorical. Encoded `<chr>`, to be mutated
- `sc`: Qualitative, categorical (social class, based on job). Encoded as numbers `12`, `3`, `45`, but these are **labels, not amounts**. To be mutated
- `carstair`: Quantitative, numeric (Carstairs deprivation score of the person's **area**; higher = more deprived)
- `smoke`: Qualitative, categorical. Encoded `<chr>`, to be mutated
- `area`: Qualitative, a code for where people live. Not a map, not a number
- `cvd`: Qualitative, categorical (cardiovascular disease, yes / no). Encoded `<chr>`, to be mutated

```{r}
#| label: point-at-a-row-B

B |> slice(1)
```

Row 1 is one person: a 50-year-old woman in social class III, who has never smoked, lives in a less deprived area (`carstair` -2.45) and doesn't have CVD.

```{r}
#| label: inspect-panel-B-data-2

skimr::skim(B)
```

On the Qual variables, panel-B data has:

- `sex` has 2 levels: "female" (4987) and "male" (3817)
- `sc` has 3 levels: `12` = classes I–II, the highest (2421), `3` = class III (3686), `45` = classes IV–V, the lowest (2697)
- `smoke` has 5 levels: "nevr" (3573), "ex" (2151), "lite" (577), "mod" (1270), "hvy" (1233)
- `cvd` has 2 levels: "no" (6389) and "yes" (2415), so about 27% have CVD
- `area` has 312 different codes

```{r}
#| label: qual-counts-B

mosaic::tally(~ sex, data = B)
mosaic::tally(~ sc, data = B)
mosaic::tally(~ smoke, data = B)
mosaic::tally(~ cvd, data = B)
```

On the Quant variables, panel-B data has:

- `age`: mean 46.2, sd 15.5, median 45, min 18, max 74
- `carstair`: mean -0.20, sd 3.27, median -0.68, min -6.23, max 12.53. Skewed right.
- Everyone in the same `area` has the same `carstair`, so it describes the neighbourhood, not the person.

There is no missing data.

```{r}
#| label: mutate-B-factors

B_mod <- B |>
  mutate(across(where(is.character), as.factor)) |>
  mutate(
    sc = fct_recode(as.factor(sc),
                    "I-II (high)" = "12",
                    "III (middle)" = "3",
                    "IV-V (low)" = "45"),
    sc = fct_relevel(sc, "I-II (high)", "III (middle)", "IV-V (low)"),
    smoke = fct_relevel(smoke, "nevr", "ex", "lite", "mod", "hvy"),
    area = as.factor(area)
  )

glimpse(B_mod)
```

`sc` gets readable names, and `smoke` goes in order from never to heavy instead of alphabetical. `age` and `carstair` stay as numbers.

One helper column: `carstair` cut into three equal groups. These are **cut-offs we chose** (equal thirds), not groups in the data.

```{r}
#| label: helper-deprivation

B_mod <- B_mod |>
  mutate(deprivation = cut_number(carstair, n = 3,
                                  labels = c("less deprived", "middle", "more deprived")))
```

All ready now.

### 2. Ask

Based on the spine (*how do class and smoke talk*), these are our questions:

Possibly can answer:

1.  Does higher social class (`sc`) go with lower deprivation (`carstair`)?
1.  Is deprivation related to CVD?
1.  Does social class affect CVD?
1.  Does smoking lower your risk of CVD? (Main Character Question)

Possibly cannot answer:

1.  Why do ex-smokers have the highest rate of CVD? Did they quit *because* they got ill? (No dates for quitting or diagnosis.)

### 3. Choose a mark

1.  Bar chart for the Qual counts (`sex`, `sc`, `smoke`, `cvd`)
1.  Boxplot for Quant + Qual (`carstair` by `sc`, `carstair` by `cvd`, `age` by `smoke`)
1.  Mosaic plot for Qual + Qual (`deprivation × sc`, `cvd × sc`, `smoke × cvd`), shaded to show surprises
1.  Chi-square test beside each mosaic: are two categories linked, or is it chance?
1.  t-test (2 groups)

### 4. Predict

1.  **Question 1:** Higher social classes will live in less deprived areas.
1.  **Question 2:** Deprivation and CVD will be inversely related.
1.  **Question 3:** Lower social classes will have more CVD.
1.  **Question 4 (Main):** Smokers will have more CVD than people who never smoked.

### 5. Make

These are the charts we used to explore the data. They don't show up in the render, but they helped us decide what to look at next.

```{r}
#| label: panelB-charts-I-want-to-see-but-not-show-1
#| eval: false
#| echo: true # Include the code, but don't compute in render

## Qual counts
B_mod |> gf_bar(~ sex)
B_mod |> gf_bar(~ sc)
B_mod |> gf_bar(~ smoke)
B_mod |> gf_bar(~ cvd)

## Quant shapes
B_mod |> gf_density(~ age)
B_mod |> gf_density(~ carstair)
```

**Question 1: does higher social class go with lower deprivation?**

```{r}
#| label: sc-deprivation

B_mod |> gf_boxplot(carstair ~ sc, fill = ~ sc) |>
  gf_labs(title = "Lower social class, more deprived neighbourhood",
          x = "Social class", y = "Carstairs score (higher = more deprived)")

vcd::structable(deprivation ~ sc, data = B_mod) |>
  vcd::mosaic(shade = TRUE, main = "Higher classes live in less deprived areas")

B_mod |> group_by(sc) |> summarise(mean_carstair = mean(carstair), n = n())
aov(carstair ~ sc, data = B_mod) |> summary()
```

Prediction was right. Class I–II averages -1.24 on `carstair`, class IV–V averages +0.65. 47% of class I–II live in the least deprived third of areas, compared to 23% of class IV–V.

**Question 2: is deprivation related to CVD?**

Our first look:

```{r}
#| label: deprivation-cvd-histogram

B_mod |>
  gf_histogram(~ carstair, fill = ~ cvd) |>
  gf_facet_wrap(~ smoke)
```

Hard to compare, because the `cvd` groups and smoking groups are such different sizes. So, a boxplot and a mosaic:

```{r}
#| label: deprivation-cvd

B_mod |> gf_boxplot(carstair ~ cvd, fill = ~ cvd) |>
  gf_labs(title = "People with CVD live in slightly more deprived areas",
          x = "CVD", y = "Carstairs score (higher = more deprived)")

vcd::structable(cvd ~ deprivation, data = B_mod) |>
  vcd::mosaic(shade = TRUE, main = "A little more CVD in more deprived areas")

mosaic::t_test(carstair ~ cvd, data = B_mod)
```

Prediction missed, in direction. It's not inverse: **more** deprivation goes with **a bit more** CVD. 25% of people in the least deprived third have CVD, compared to 29% in the most deprived third. The gap is small, but the t-test says it's real (p far below 0.05).

**Question 3: does social class affect CVD?**

```{r}
#| label: sc-cvd

vcd::structable(cvd ~ sc, data = B_mod) |>
  vcd::mosaic(shade = TRUE, main = "CVD barely changes across social class")

B_mod |> group_by(sc) |> summarise(pct_cvd = mean(cvd == "yes"), n = n())
mosaic::xchisq.test(mosaic::tally(cvd ~ sc, data = B_mod))
```

Only a small difference: 25.7% (I–II), 27.8% (III), 28.5% (IV–V). The chi-square p-value is about 0.06, just above 0.05, so we can't say class and CVD are linked here.

Class does talk to smoking, though:

```{r}
#| label: sc-smoke

vcd::structable(smoke ~ sc, data = B_mod) |>
  vcd::mosaic(shade = TRUE, main = "Lower classes smoke more heavily")
```

48% of class I–II have never smoked, compared to 35% of class IV–V. Heavy smokers: 9% vs 19%.

**Question 4 (Main): does smoking lower your risk of CVD?**

Our mosaics and test:

```{r}
#| label: smoke-cvd

vcd::structable(smoke ~ cvd, data = B_mod) |> vcd::mosaic()
vcd::structable(smoke ~ cvd, data = B_mod) |> vcd::mosaic(type = "expected")
vcd::structable(smoke ~ cvd, data = B_mod) |>
  vcd::mosaic(shade = TRUE, main = "Ex-smokers stand out with the most CVD")

B_mod |> group_by(smoke) |> summarise(pct_cvd = mean(cvd == "yes"), n = n())
mosaic::xchisq.test(mosaic::tally(smoke ~ cvd, data = B_mod))
```

Prediction missed. Current smokers (lite, mod, hvy: 24–26%) have about the same CVD rate as people who never smoked (24%). The group that stands out is **ex-smokers, at 36%**. Why?

```{r}
#| label: smoke-age

B_mod |> gf_boxplot(age ~ smoke, fill = ~ smoke) |>
  gf_labs(title = "Ex-smokers are older than everyone else", x = NULL, y = "Age")

B_mod |> group_by(smoke) |> summarise(mean_age = mean(age))
```

Ex-smokers are older (average 52, compared to 42–45 for the other groups), and CVD goes up a lot with age. So age is probably doing a lot of the work here.

### 6. Say

1.  Class and place talk: higher social classes live in less deprived areas (class I–II average `carstair` -1.24 vs +0.65 for class IV–V, p far below 0.05).
1.  More deprived areas have slightly more CVD (25% vs 29%), the opposite direction from our prediction. Small, but real.
1.  Social class on its own barely changes CVD (p ≈ 0.06). Class shows up in *where people live* and *how much they smoke* more than in CVD directly.
1.  Smoking does **not** lower your risk. Current smokers have about the same CVD rate as never-smokers, and ex-smokers have the most (36%), but they're also the oldest group.

### 7. Limit

- One snapshot from 1998. Links, not causes.
- We don't know when people quit smoking or when CVD was diagnosed, so some ex-smokers may have quit *because* they got ill.
- `smoke` is in groups, and we don't know what "lite" or "hvy" actually means in cigarettes.
- `carstair` describes the area, not the person: a well-off person in a deprived area gets the area's score.
- `sc` is based on job type, which leaves out people without a job and says nothing about income.
- `cvd` is just yes / no, with no type or severity.
- Age affects almost everything here, and we haven't fully separated it out.
- The deprivation groups are cut-offs we chose (equal thirds).
- Kit table we refused to abuse: `sc` and `area` were never averaged or treated as numbers, and `area` was never used as a map.

# Conclusion

- **The spine so far:** Bodies carry risk differently by sex (A), class talks to *where people live* and *how much they smoke* more than to heart disease directly (B), and smoking keeps refusing to behave the way we predicted in both panels.
- **The questions we will defend in ninety seconds:**
    - Panel A: Do smokers and non-smokers carry cholesterol differently, and why do male smokers look *healthier*?
    - Panel B: Does smoking lower your risk of CVD? (No: ex-smokers have the most CVD, and they're also the oldest.)
- **What we will not claim:** that smoking is good for cholesterol or the heart, that social class causes CVD, or that anything here causes heart disease.
- **Which lines were human, which were machine:**
    - Human: the questions in both panels, the predictions, the first HDL/LDL histograms, the sys/dia scatter, using `facet_wrap()` by sex, spotting the "how many cigarettes" limit, the helper columns and cut-offs, the violins and `tally()` counts, the cholesterol-by-sex histogram and t-test from class, the `carstair` histogram faceted by `smoke`, and the `smoke × cvd` mosaics and chi-square test.
    - Machine (AI-assisted): the `pivot_longer()` combined chart,, formatting `index.qmd` into the template layout, merging drafts into one file, formatting Panel B, the deprivation groups (`cut_number()`), the age check for ex-smokers, and checking the numbers in both panels.
- **Limits:** one snapshot each, smoking is yes / no only in A and in loose groups in B, only 36 female smokers in A, age tangled up with smoking in B, `carstair` describes the area not the person, unknown sampling.
- **Gift for the steal-this strip:** `pivot_longer()` to put two measures side by side in one faceted chart, helper columns with cut-offs stated out loud, and `fct_relevel()` / `fct_recode()` to put categories in a sensible order with readable names.